% !TEX root = master.tex

\chapter*{Acknowledgements}

I owe an enormous debt of gratitude to all of the project partners.  Tahoe National Forest, Terri Walsh, Marilyn Tierney, and Alan Doerr from the Tahoe National Forest, and Becky Estes and Hugh Safford, from the U.S. Forest Service Region 5 Ecology Group, provided data, insight into the local ecology, and invaluable expert opinion throughout the process. Thank you for all of your enthusiasm, for your critical analysis of intermediate results, and for your help in learning the ecology of an amazing ecosystem. We made a lot of decisions together and your careful thoughts about each one were critical. I learned so much from all of you---thank you for being my colleagues during this study! In addition, thanks are due to Kevin McGarigal, who obtained funding for this project and hired me to work on it nearly four years ago. Thank you for giving so much of your time to steadily work towards completion. Your high expectations kept me on track and led to better research. To Eduard Ene, the programmer I never met, without whom none of the simulation or analysis would have happened. Thanks for all the bug fixes.

To my committee, all of whom invested much more time into this process than I (naively) initially promised they would have to commit. Jack, thank you for keeping me focused on the big picture and providing reassurance when I needed it. Brad, thank you for your careful evaluation of my strategy for analyzing a future range of variability, and for being unfailingly supportive and honest. Becky, I can't thank you enough for your partnership through the HRV component of the project, your extensive, insightful, and thoughtful comments on the FRV component and on many drafts of this thesis, and for mentoring me throughout the entire process. I loved working with you and hope I'm lucky enough to do so again professionally.

I was fortunate during my time at UMass to develop relationships with faculty outside of my committee. Allison and Paige, I'm so glad such wonderful women are in this department. I really appreciate that you always made time and space for me, offering advice and support, despite how busy you are. Brian, thank you for believing in me and recruiting me before anyone else did! I will always appreciate your friendship, support, and mentorship, and your fairness and good advice. I'm so glad you had to use a student office as your lab four years ago.

Partway through my degree, I was hired by the North Atlantic LCC, an experience which has given me invaluable insight on the flip side of projects similar to mine. This thesis is better because of my professional experiences with the North Atlantic LCC. I am incredibly fortunate to have more friends and colleagues at the U.S. Fish and Wildlife Service office than I can list here, but one deserves an extended thank you. Scott, for hiring me and giving me so many opportunities to grow as a scientist in a nonacademic environment, I am deeply grateful. Thank you also for your steadfast support and encouragement, and for consistently demonstrating kindness and selflessness. You are a role model to me both personally and professionally.

I don't know how I would have accomplished this research without my ECO friends. Blake, Rebekah, Melanie, and Kendra, I am so grateful for your friendship and support. Knowing I have people like you to turn to when I need someone to listen or offer advice is invaluable. Thanks for the adventures, the drinks, the food, and the conversation. To my parents, thank you for your unending support, for listening to all my frustrations, for making sure I always know that I am safe and loved, and for being genuinely interested in my research. Thanks to Tilly, for being silly and unfailingly adorable, and making me smile and laugh so much. Last and most importantly, Lee, thank you for keeping my spirits up, for so much editing, for solving some of the trickier statistics questions, for helping me learn how to be a scientist, and for being the best. 

