% !TEX root = master.tex

\chapter{Historical Range of Variability}
\label{ch:hrv}
\setcitestyle{notesep={:},aysep={}}

In this chapter I present the methods, results, and discussion for my analysis of the historical range of variability (HRV)in disturbance and forest structure in the Yuba River Watershed in the northern Sierra Nevada. The historical period defined for this project was 1550 to 1850. Unfortunately, I did not have consistent and complete empirical data on wildfire and vegetation growth and pattern during this period. Instead, I used simulations to incorporate the data that do exist, generate new datasets of otherwise unobservable landscape trajectories, and ultimately quantify the HRV for my study area. Many landscape disturbance and succession models exist, with different input requirements, purposes, intended geographic applicability, etc. Typically, each timestep of the model produces a set of GIS layers, which can then be analyzed to quantify trajectories and patterns; in other words, to describe the range of variability. In this study I used the Rocky Mountain Landscape Simulator (\textsc{RMLands}). Originally developed for use in the Rocky Mountains, an early phase of this project was to adapt \textsc{RMLands} to the Sierra Nevada. \textsc{RMLands} is disturbance and succession model software that is stochastic, spatially explicit, and raster-based. The model was parameterized to simulate passive management of fire and vegetation. That is, I did not simulate vegetation treatments, nor did I attempt to emulate additional fire suppression efforts.

\input{Ch2_S1_Methods}
\input{Ch2_S2_Results}
\input{Ch2_S3_Discussion}